Deciding Bitvector Termination with Program Synthesis

### 2014-07-03 papers/termination/kalashnikov.tex
revised more of the kalashnikov section

\iffalse
\section{Introduction}

Program synthesis is the mechanised construction of software that provably
satisfies a given specification.  Synthesis tools promise to relieve the
programmer from thinking about \emph{how} the problem is to be solved;
instead, the programmer only provides a compact description of \emph{what}
is to be achieved.  Foundational research in this area has been
exceptionally fruitful, beginning with Alonzo Church's work on the
\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
Algorithmic approaches to the problem have frequently been connected to
automated theorem proving~\cite{manna-waldinger,proof-planning}.
Recent developments include an application of Craig interpolation
to synthesis~\cite{synth-interpolants}.

In this paper, we present a new method that brings the promise of
automating programming closer to application.  We make pragmatic
restrictions to the general synthesis problem in order to arrive at an
effective solution.  To this end, we focus on the problem of synthesising
loop-free programs over machine integers and floating-point numbers from a
specification given as program fragment.  Our exploration of the program
space ensures that our programs are minimal in size.

Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
which is the closest work to ours.  {\sc Brahma} also focuses on
synthesising loop-free machine-integer programs.  In~\cite{brahma}, the
authors characterise the program synthesis problem as a second-order
$\exists \forall$ formula and use a variation of the CEGIS algorithm,
introduced in~\cite{lezama-thesis}, for solving
these formulae.  This algorithm reduces the satisfiability of the $\exists
\forall$ formula to the satisfiability of several first-order formulae, each
of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
takes as input a specification (given as an SMT formula) and a library of
components (one component is more or less a single machine instruction).  It
then finds a permutation of the component library that satisfies the
specification.  In the
event that the library does not contain enough components to synthesise a
correct program, the user is told that the library is too small and
must add more components until the synthesis succeeds.  Each round of
synthesis amounts to solving an $\exists \forall$ formula that is quadratic
in the size of the component library.

We require the user to supply a program specification written in C, but then
ask nothing more of her.  We parametrise the space of programs in such a way
that it can be explored automatically, rather than asking a human for hints.
Our program encoding generates an $\exists
\forall$ formula that is \emph{linear} in the length of the shortest program
satisfying the user's specification.  This formula is then checked for satisfiability
using the CEGIS algorithm.  In contrast, the program generated by {\sc
brahma} can be no longer than the size of the library, as the synthesis
outcome is a permutation of the component library.  Our synthesis
formulae are therefore asymptotically smaller than {\sc brahma}'s.

\paragraph{Related Work}

Another successful approach to program synthesis has been \emph{program
sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch,modular-sketch}.  This method
requires the programmer to provide a \emph{program sketch}, written in a
C-like language, together with either a specification (written in the same
language) or a suite of unit tests.  The sketch is a skeleton program
containing much of the high-level structure of the final program, but with
several ``holes'', which {\sc sketch} fills in to meet the specification.
Despite the many similarities, our problem domains are quite different:
sketching is able to make use of human insight to synthesise fairly large
programs, whereas we are optimised for synthesising small (but intricate)
programs without guidance.  A sketching-like approach to optimising
fixed point arithmetic code was described in~\cite{eldib-wang}.  This
work is similar in goal to our floating-point work, but is more restricted
in scope in that it only tries to minimise register widths for embedded
systems using fixed point arithmetic.

Program synthesis is closely related to
superoptimisation, which is the problem of taking a
segment of code, usually written in C, and finding a minimal program which
has the same functionality.  The {\sc aha}~\cite{aha-paper} tool, the GNU
superoptimiser~\cite{gnu-superoptimiser} and TOAST~\cite{toast} all perform superoptimisation of C
programs by enumerating possible programs and testing them on suitable
inputs.  This search can be compiled down to native code using an optimising
compiler, so for short programs the method is extremely fast.  However, this
method is impractical for synthesising longer programs because of the
exponential blowup of the search space.  We incorporate the core idea of
this technique by using explicit-state model checking in parallel with
symbolic model checking.

\paragraph{Technical Contributions}
Our approach generates an $\exists \forall$ formula that is linear
in the size of the synthesised program.  This improves over the
encoding of~\cite{brahma}, which generates a $\exists \forall$ formula
that is quadratic in the size of the synthesised program.

We introduce a novel parametrisation of the programming language
used to express our synthesised programs.
This parametrisation allows us to efficiently explore the program space
without relying on human guidance and also ensures that our programs
are of minimal length.

Our tool is the first we are aware of that is able to effectively
synthesise floating-point programs.  We demonstrate this by
synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
a specification.

\paragraph{Outline} We describe the abstract synthesis algorithm in
Sec.~\ref{sec:abstract-algorithm} and give details of our new algorithm in
Sec.~\ref{sec:concrete-algorithm}.  We use a fragment of the C programming language
as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
summarise the results of our experiments in Sec.~\ref{sec:experiments}.
\fi

\iffalse
\subsection{The Basic Synthesis Algorithm}
\label{sec:abstract-algorithm}

\subsubsection{The Synthesis Formula}
\label{sec:abstract-formula}

Our task is to find a program that satisfies some specification.  We can formalise this
notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
relate inputs and outputs, and programs $P$ map an input to an output:
%
$$ \sigma \subseteq I \times O $$
$$ P : I \rightarrow O$$

The synthesis problem is then to determine the validity of the following
second-order formula, and if so, to find a suitable witness program $P$:
%
$$\exists P .\, \forall x \in I.\, \sigma(x, P(x))$$

Depending on the logic needed to express this formula the synthesis problem
may be decidable, semi-decidable or undecidable.  For many interesting
logics, checking the validity of a second-order formula is undecidable. 
Even for logics in which the problem is decidable, the quantifier
alternation means that we may not have an efficient decision procedure for
the formula above, despite efficient methods existing for problems
with only one quantifier.  In the case of propositional
satisfiability, we have SAT solvers which are efficient-in-practice at
solving queries with a single quantifier.  The corresponding problem with
one level of quantifier alternation (2-QBF) is complete for $NP^{NP}$, and
current 2-QBF solvers are much less efficient in practice than SAT solvers.

\subsubsection{CEGIS}

\fi

In this section we present a solver for the synthesis fragment. An
important characteristic of our solver is that is models bit accurate
semantics. This enables the precise treatment of bit-wise operations and
arithmetic modulo fixed widths.
% 
%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
%have a program synthesiser kicking around, we present the design of one synthesiser
%that can be used to solve the problem.
%We start by recalling the synthesis formula:
% $$\exists P, x . \forall y . \sigma(P, x, y)$$
We start by recalling the synthesis fragment:
 \[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
 \]

where $P$ ranges over sets, while $x$ and $y$ range over ground terms (we have dropped
 the vector symbols for brevity reasons).

\subsection{\newC}
\label{sec:logic}
In order to obtain bit-level accuracy, the logic we use to express the synthesis formula is a subset of C that we call \newC.  
The characteristic property of a
\newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

The proofs found by our solver are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
\todo{update the figure and make it look more like a grammar}

As the quantifier free first-order fragment of the logic is efficiently decidable and
$P$ is expressible as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula.

@@

\iffalse
\subsection{The Concrete Algorithm for Bitvector Programs}
\label{sec:concrete-algorithm}

One area in which program synthesis can shine is in producing very small,
intricate programs that manipulate bitvectors.  An example of such a program
is given in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
as input and clears every bit except for the least significant bit that was set.
Even though this program is extremely short, it is fairly difficult for a human
to see what it does.  It is even more difficult for a human to come up with such
minimal code.  The program is so concise because it
takes advantage of the low-level details of the machine, such as the fact that
signed integers are stored in two's complement form.

\begin{figure*}
\centering
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C]
int isolate_lsb(int x) {
  return x & -x;
}
 \end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}

Example:

\hrule

\begin{tabular}{llcccccccc}
 x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
 -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
 x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
\end{tabular}
\end{minipage}

 \caption{A tricky bitvector program}
  \label{fig:bitvector-program}
\end{figure*}

To synthesise tricky bitvector programs like this, it is natural for us to
work in the logic of quantifier-free propositional formulae and to use a
propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
However we propose a slightly different tack, which is to use a decidable
fragment of C as a ``high level'' logic.
\fi

%% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
%% \sigma$ and $P$ in \newC, then ensure that we can express the validity of the
%% synthesis formula as a safety property of the resulting \newC program.

%% Our encoding is the following:
%% %
%% \begin{itemize}
%%  \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
%%  \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which is written in \newC as the type \verb|int[M]|.
%%  \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
%%  \verb|int check(int in[N], int out[M])|. This function is the only component supplied
%%  by the user.
%%  \item $P$ is written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
%%  have type $I \rightarrow O$ and
%%  are represented in \newC as objects of type \verb|prog_t|, shown in Fig.~\ref{fig:c-l-encoding}.
%%  \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
%%  this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
%%  \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an output parameter.
%% \end{itemize}

@@

%% The exact details of how we encode an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
%% We must now express the {\sc synth} and {\sc verif} formulae as safety properties
%% of \newC programs, which is given in Fig.~\ref{fig:c-synthverif}.

%% \begin{figure}
%% \centering
%% \begin{minipage}[t]{.45\textwidth}
%% \begin{lstlisting}[language=C++]
%% void synth() {
%%   prog_t p = nondet();
%%   int in[N], out[M];

%%   assume(wellformed(p));

%%   in = test1;
%%   exec(p, in, out);
%%   assume(check(in, out));
%%   ...
%%   in = testN;
%%   exec(p, in, out);
%%   assume(check(in, out));

%%   assert(false);
%% }
%% \end{lstlisting}
%% %\end{minipage}
%% %\hfill
%% %\begin{minipage}[t]{.45\linewidth}
%% \begin{lstlisting}[language=C++]
%% void verif(prog_t p) {
%%   int in[N] = nondet();
%%   int out[M];

%%   exec(p, in, out);
%%   assert(check(in, out));
%% }
%% \end{lstlisting}
%% \end{minipage}

%%  \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program.}
%%  \label{fig:c-synthverif}
%% \end{figure}

@@

\node [file] (synth) {\sc synth.c};
\path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
\path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
\path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};

\path (tests.east)+(2.0, -0.25) node [process] (merged) {merge};

\path (merged.east)+(2.0, 1.0) node [process] (cbmc) {\sc cbmc};
\path (merged.east)+(2.0, -1.0) node [process] (gcc) {\sc gcc};

\path (cbmc.east)+(2.5, -1.0) node [file] (out) {candidate program};

\path [line] (synth) -- (merged);
\path [line] (tests) -- (merged);
\path [line] (interpreter) -- (merged);
\path [line] (spec) -- (merged);

\path [line] (merged) -- (cbmc);
\path [line] (merged) -- (gcc);

\path [line] (cbmc) -- (out);
\path [line] (gcc) -- (out);

@@

%\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
%% \label{sec:encode-l}

%% \begin{figure}
%% \begin{lstlisting}[language=c++,mathescape]
%% typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
%% typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
%% typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand

%% struct prog_t {
%%   op_t ops[$l$];          // The opcodes
%%   param_t params[$l$*2];  // The operands
%%   word_t consts[$c$];     // The program constants
%% }
%% \end{lstlisting}

%%  \caption{The \newC structure we use to encode an $\mathcal{L}$ program
%%   \label{fig:c-l-encoding}}
%% \end{figure}

%% The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
%% The \verb|prog_t| structure encodes a program, which is a sequence of instructions.
%% The parameter $a$ is the number of arguments the program takes.
%% The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
%% right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
%% a program argument or the result of a previous instruction, and its value
%% is determined at runtime as follows:
%% \[
%%  val(x) = \begin{cases}
%%            x < a & \text{the } x^{\text{th}} \text{ program argument} \\
%%            a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
%%            x \geq a + c & \text{the result of instruction } (x - a - c)
%%           \end{cases}
%% \]

%% A program is well formed if
%% no operand refers to the result of an instruction that has not been computed yet, and if
%% each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
%% for each instruction.  It should be noted that this requires a linear number of well-formedness
%% constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.

@@

In order to search the space of candidate programs, we parametrise
the language~$\mathcal{L}$, inducing a lattice of progressively
more expressive languages.  We start by attempting to synthesise
a program at the lowest point on this lattice and increase the
parameters of~$\mathcal{L}$ until we reach a point at which
the synthesis succeeds.

@@

---
We require the user to supply a program specification written in C, but then
ask nothing more of her.  We parametrise the space of programs in such a way
that it can be explored automatically, rather than asking a human for hints.
Our program encoding generates an $\exists
\forall$ formula that is \emph{linear} in the length of the shortest program
satisfying the user's specification.  This formula is then checked for satisfiability
using the CEGIS algorithm.
---

We introduce a novel parametrisation of the programming language
used to express our synthesised programs.
This parametrisation allows us to efficiently explore the program space
without relying on human guidance and also ensures that our programs
are of minimal length.

Our tool is the first we are aware of that is able to effectively
synthesise floating-point programs.  We demonstrate this by
synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
a specification.
------

We consider the following parameters:
\begin{itemize}
\item{Program Length: $l$}
%% The first parameter we introduce is program length, denoted by $l$.
%% At each iteration we synthesise programs of length exactly $l$.
%% We start with $l = 1$ and increment $l$ whenever we determine
%% that no program of length $l$ can satisfy the specification.  When we do
%% successfully synthesise a program, we are \emph{guaranteed that it
%% is of minimal length} since we have previously established that no
%% shorter program is correct.
\item{Word Width: $w$}
%% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
%% has its own set of parameters.  The only relevant parameter is
%% the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
%% in each internal register and immediate constant.  This parameter is denoted by
%% $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
%% polynomially with $w$, since each intermediate C variable corresponds
%% to $w$ propositional variables.

%% It is often the case that a program which satisfies the specification
%% on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
%% specification when run on a machine with $w > k$.  For example, the program
%% in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
%% This is true irrespective of the word size of the machine it is run on -- it will
%% isolate the least-significant bit of an 8-bit word just as well as it will a
%% 32-bit word.  An often successful strategy is to synthesise a program for an
%% $\mathcal{L}$-machine with a small word size and then to check whether the
%% same program is correct when run on an $\mathcal{L}$-machine with a
%% full-sized word.

%% The only wrinkle here is that we will sometimes synthesise a program containing
%% constants.  If we have synthesised a program with $w=k$,
%% the constants in the program will be $k$-bits wide.  To extend the program
%% to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
%% numbers from $k$-bit ones.  We have several strategies for this and
%% just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
%% $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
%% and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.

%% \begin{figure}

%% \centering
%% \begin{minipage}[t]{.45\textwidth}
%% \begin{eqnarray*}
%%  \bv{m}{m} & \rightarrow & \bv{n}{n} \\
%%  \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
%%  \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
%% \end{eqnarray*}
%% \end{minipage}
%% \begin{minipage}[t]{.45\textwidth}
%% \begin{eqnarray*}
%%  \bv{x}{m} & \rightarrow & \bv{x}{n} \\
%%  \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
%%  \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
%% \end{eqnarray*}
%% \end{minipage}

%% \caption{Rules for extending
%% an $m$-bit wide number to an $n$-bit wide one.
%%  \label{fig:generalize}}
%% \end{figure}

%% Sometimes a program will be correct for some particular word width $w$, but is
%% not correct for $w' > w$ even if the constants are replaced with appropriate ones.
%% When we detect this situation, we increase $w$ and continue synthesising.

\item{Number of Constants: $c$}
%% Instructions in $\mathcal{L}$ take either one or two operands.
%% Since any instruction whose operands are all constants can always be
%% eliminated (since its result is a constant), we know that a loop-free program
%% of minimal length will not contain any instructions with two constant
%% operands.  Therefore the number of constants that can appear in
%% a minimal program of length $l$ is at most $l$.  By minimising the number
%% of constants appearing in a program, we are able to use a particularly
%% efficient program encoding that speeds up the synthesis procedure
%% substantially.  The number of constants used in a program is the parameter $c$.

%% $\mathcal{L}$ is an SSA, three-address instruction set\footnote{
%% We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
%% the programs to be smaller and synthesis to be faster as a result.  We saw
%% the opposite effect -- the more complex interpreter led to much slower synthesis.
%% }.  Destination registers
%% are implicit and a fresh register exists for each instruction to write its
%% output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
%% opcode and two operands, where each operand is either a register (i.e., a program argument
%% or the result of a previous instruction), or an immediate constant.

%% In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
%% of instruction types in $\mathcal{L}$.  Each operand can be encoded using
%% $\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
%% bit to specify whether the operand is a register name or an immediate constant.
%% One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
%% For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
%% the entire program.

%% If we instead limit the number of constants that can appear in the program,
%% our operands can be encoded using fewer bits.  For an $n$-instruction program
%% using $c$ constants and taking $a$ arguments as inputs, each operand can refer
%% to a program argument, the result of a previous instruction or a constant.
%% This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
%% can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
%% needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.

%% We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
%% For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
%% machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
%% Using the second encoding, each operand can be represented using
%% $\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
%% This is a substantial reduction in size and when the desired program requires
%% only few constants this can lead to a very significant speed up.

%% As with program length, we progressively increase the number of constants in
%% our program.  We start by trying to synthesise a program with no constants,
%% then if that fails we attempt to synthesise using one constant and so on until we
%% reach $c = l$.

\end{itemize}
\subsubsection{Searching the Program Space}

The key to our automation approach is to come up with a sensible way in which to
adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
After each round of {\sc synth}, we may need to adjust the parameters.  The
logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.

Whenever {\sc synth} fails, we consider which parameter might have caused the
failure.  There are two possibilities: either the program length $l$ was too small,
or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
try another round of synthesis, but allowing ourselves an extra program constant.
If $c = l$, there is no point in increasing $c$ any further.  This is because
no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
have to be at least one instruction with two constant operands.  This
instruction could be removed (at the expense of adding its result as
a constant), contradicting the assumed minimality of the program.  So
if $c = l$, we set $c$ to 0 and increment $l$, before attempting
synthesis again.

If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
program that is correct for some inputs but incorrect on at least
one input.  However, it may be the case that the candidate program
is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
with a small word size.  For example, we may have synthesised a
program which is correct for all 8-bit inputs, but incorrect for
some 32-bit input.  If this is the case (which we can determine
by running the candidate program through {\sc verif} using the smaller
word size), we may be able to produce a correct program for
the full $\mathcal{L}$-machine by using the constant extension rules
shown in Fig.~\ref{fig:generalize}.  If constant generalization
is able to find a correct program, we are done.  Otherwise,
we need to increase the word width of the $\mathcal{L}$-machine
we are currently synthesising for.

%% \begin{figure}[t]
%% \centering
%% \begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
%% \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
%%     text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
%% \tikzstyle{block} = [rectangle, draw, fill=blue!20, 
%%     text width=5em, text centered, rounded corners, minimum height=4em]
%% \tikzstyle{line} = [draw, -latex']
%% \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
%%     minimum height=2em]

%%  \node [decision] (synthsucc) {{\sc Synth} succeeds?};

%%  \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
%%  \node [decision, right of=verif] (ck) {$c < l$?};

%%  \node [block, below of=verif, node distance=3cm]   (done) {Done!};
%%  \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};

%%  \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
%%  \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};

%%  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
%%  \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};

%%  \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};

%%  \path [line] (synthsucc) -- node [left] {Yes} (verif);
%%  \path [line] (synthsucc) -| node [above, near start] {No} (ck);

%%  \path [line] (verif) -- node [left] {Yes} (done);
%%  \path [line] (verif) -- node [above, near start] {No} (verifw);

%%  \path [line] (ck) -- node [left] {Yes} (incc);
%%  \path [line] (ck) -| node  [above, near start]  {No} (incl);

%%  \path [line] (verifw) -- node [right] {Yes} (gen);
%%  \path [line] (verifw) -- node [above] {No} (iterate);

>>>>

%\todo{add related work for synthesis.}

In this section we present a solver for the synthesis fragment:
 \[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
 \]
%
where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as mentioned before, for brevity reasons we drop
 the vector symbol and write  $x$ for $\vec{x}$).

By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ as a specification function
that  returns true iff $P$ computes appropriate outputs for the inputs $x$ and $y$, we can   
regard the satisfaction problem for the synthesis fragment as \emph{program synthesis}, i.e.
the mechanised construction of software that provably satisfies a given specification.
Research in the area of program synthesis has been particularly fruitful beginning with Alonzo Church's work on the
\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and continuing with works such as 
{\sc Brahma}~\cite{brahma} and program sketching~\cite{lezama-thesis,sketch,modular-sketch}.

% 
%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
%have a program synthesiser kicking around, we present the design of one synthesiser
%that can be used to solve the problem.
%We start by recalling the synthesis formula:
% $$\exists P, x . \forall y . \sigma(P, x, y)$$

%% Our program encoding generates an $\exists\forall$ formula that is \emph{linear} in the length of the shortest program
%% satisfying the user's specification.  This formula is then checked for satisfiability
%% using the CEGIS algorithm.

In the context of (non-)termination analysis, a witnesses to the satisfiability of a formula in the synthesis fragment 
is a (non-)termination proof. 
As mentioned earlier in the paper, our analysis must ensure precise treatment of bit-wise operations and arithmetic modulo fixed widths
%Consequently, a compulsory characteristic of our solver is that it must 
by modelling bit accurate semantics.
The synthesiser must therefore be able to compute (non-termination) proofs 
%loop-free programs 
over machine integers and floating-point numbers from the specifications detailed in Section~\ref{sec:second.order}.

%\subsection{\newC}
%\label{sec:logic}
\subsection{The Synthesis Specification}
Our (non-)termination specifications are given as program fragments
%More specifically, the logic we use to express the synthesis formula
in a subset of C that we call \newC.  
The characteristic property of a \newC  program is that safety can be decided for
it using a single query to a Bounded Model Checker.  A \newC program is
just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
\todo{update the figure and make it look more like a grammar.}

\subsection{The Synthesis Algorithm}
As the quantifier free first-order fragment of the logic is efficiently decidable and
$P$ is expressible as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula.

@@

@@

@@

\node [file] (spec) {\sc Specification};
%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};

%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};

\path (spec.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
\path (spec.east)+(2.0, 0.0) node [process] (gcc) {\sc gcc};
\path (spec.east)+(2.0, -1.5) node [process] (gp) {\sc GP};

\path (gcc.east)+(2.5, 0.0) node [file] (out) {Proof};

%\path [line] (spec) -- (merged);

\path [line] (spec) -- (cbmc);
\path [line] (spec) -- (gcc);
\path [line] (spec) -- (gp);

\path [line] (cbmc) -- (out);
\path [line] (gcc) -- (out);
\path [line] (gp) -- (out);

@@

@@

In order to search the space of candidate programs, we parametrise
the language~$\mathcal{L}$ by program length, word width and number of constants,
inducing a lattice of progressively
more expressive languages.  We start by attempting to synthesise
a program at the lowest point on this lattice and increase the
parameters of~$\mathcal{L}$ until we reach a point at which
the synthesis succeeds.

@@

%% We introduce a novel parametrisation of the programming language
%% used to express our synthesised programs.
%% This parametrisation allows us to efficiently explore the program space
%% without relying on human guidance and also ensures that our programs
%% are of minimal length.

%% Our tool is the first we are aware of that is able to effectively
%% synthesise floating-point programs. We demonstrate this by
%% synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
%% a specification.

%% ---Our exploration of the program space ensures that our programs are minimal in size.
%% --- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.

%% We consider the following parameters:
%% \begin{itemize}
%% \item{Program Length: $l$}
%% The first parameter we introduce is program length, denoted by $l$.
%% At each iteration we synthesise programs of length exactly $l$.
%% We start with $l = 1$ and increment $l$ whenever we determine
%% that no program of length $l$ can satisfy the specification.  When we do
%% successfully synthesise a program, we are \emph{guaranteed that it
%% is of minimal length} since we have previously established that no
%% shorter program is correct.
%\item{Word Width: $w$}
%% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
%% has its own set of parameters.  The only relevant parameter is
%% the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
%% in each internal register and immediate constant.  This parameter is denoted by
%% $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
%% polynomially with $w$, since each intermediate C variable corresponds
%% to $w$ propositional variables.

%\item{Number of Constants: $c$}
%% Instructions in $\mathcal{L}$ take either one or two operands.
%% Since any instruction whose operands are all constants can always be
%% eliminated (since its result is a constant), we know that a loop-free program
%% of minimal length will not contain any instructions with two constant
%% operands.  Therefore the number of constants that can appear in
%% a minimal program of length $l$ is at most $l$.  By minimising the number
%% of constants appearing in a program, we are able to use a particularly
%% efficient program encoding that speeds up the synthesis procedure
%% substantially.  The number of constants used in a program is the parameter $c$.

%\end{itemize}
%\subsubsection{Searching the Program Space}

The key to our automation approach is to come up with a sensible way in which to
adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
After each round of {\sc synth}, we may need to adjust the parameters.  
%The
%logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
Whenever {\sc synth} fails, we consider which parameter might have caused the
failure.  
%% There are two possibilities: either the program length $l$ was too small,
%% or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
%% try another round of synthesis, but allowing ourselves an extra program constant.
%% If $c = l$, there is no point in increasing $c$ any further.  This is because
%% no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
%% have to be at least one instruction with two constant operands.  This
%% instruction could be removed (at the expense of adding its result as
%% a constant), contradicting the assumed minimality of the program.  So
%% if $c = l$, we set $c$ to 0 and increment $l$, before attempting
%% synthesis again.

%% If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
%% program that is correct for some inputs but incorrect on at least
%% one input.  However, it may be the case that the candidate program
%% is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
%% with a small word size.  For example, we may have synthesised a
%% program which is correct for all 8-bit inputs, but incorrect for
%% some 32-bit input.  If this is the case (which we can determine
%% by running the candidate program through {\sc verif} using the smaller
%% word size), we may be able to produce a correct program for
%% the full $\mathcal{L}$-machine by using the constant extension rules
%% shown in Fig.~\ref{fig:generalize}.  If constant generalization
%% is able to find a correct program, we are done.  Otherwise,
%% we need to increase the word width of the $\mathcal{L}$-machine
%% we are currently synthesising for.

%% \section{Conclusion}

%% By expressing the program synthesis problem as a safety property for a
%% program interpreter, we have been able to harness the power of
%% state-of-the-art program analysis tools and reuse them in a new problem
%% domain.  We have implemented our algorithm as a freely downloadable tool
%% whose performance compares favourably to a recent program synthesiser. 
%% Finally, we have taken advantage of the expressiveness of our specification
%% language to make an initial step towards practical synthesis of
%% floating-point programs.
